---
title: "Use of whomapper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use of whomapper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9.44, 
  fig.height = 5.76,
  warning = FALSE, 
  message = FALSE
)
```

## Introduction

{whomapper} is a R package internal to WHO, intended to house shapefiles and produce WHO legal maps. While this package can produce legal maps, please note that **the user of this package is responsible for ensuring that any maps they produce fulfill WHO legal standards**. The WHO map SOP can be accessed [here](http://gamapserver.who.int/gho/gis/training/DMF_GIS2010_2_SOPSforWHOMaps.pdf){target="_blank"}.

We start by loading admin 0 shapefiles, and WHO vaccine data for visualisation by country:

```{r setup}
pacman::p_load(tidyverse, whomapper, sf, lubridate)

## we use WHO vaccine data to visualise on a map
vaccine_data <- whomapper:::vaccine_data

## pull in shapefiles for admin 0
## We use the offline shapefiles here, but better practice may be getting online files
sfs <- whomapper::pull_sfs(adm_level = 0, query_server = FALSE)

```


## Producing basic maps at admin 0 level

This package has been set up to allow for flexible use of `ggplot2` for mapping. It's highly recommended that `tidyverse` is installed before using this package to manipulate maps. The basic workflow for using this package is:

1. Read in shapefiles (packaged with {whomapper})
2. Merge data of interest with the `adm0` list element of the shapefiles (the polygons)
3. Begin a standard ggplot:
    1. Plot shapefiles with `geom_sf_who_poly()`
    2. Define legend aesthetics with `scale_...()`
    3. Define labels for the plot
4. call `who_map_pipeline()` to define disputed areas, theme, legal text etc.
5. Save the plot using `who_map_save()` - This ensures that dimensions are correct for global maps


Some basic tips for using this package are:

1. Use the `geom_sf_who_poly()` custom geom to plot shapefile **polygons** with default settings (i.e. no border)
2. Legend aesthetics can be controlled as normal with ggplot - here this should be done before calling `who_map_pipeline()` (as in the below examples).
3. The caption, x-axis, and y-axis labels cannot be defined - only the title and subtitle are available.
4. Use `who_map_col()` to define some colours that must be explicit. For example, where there is no data, use `who_map_col("no_data")`

Note that this package can be used to plot maps with continuous, binned, and categorical data. Some examples are shown below.


```{r basic_map}

ggplot() +
  geom_sf_who_poly(data = sfs$adm0, aes(fill = who_region)) +
  scale_fill_brewer(palette = "Set3", 
                    na.value = who_map_col("no_data"),
                    name = "WHO Region") +
  labs(title ="World map by WHO region", 
       subtitle = str_glue("as of {format(Sys.Date(), '%d %b %y')}")) +
  who_map_pipeline(sf = sfs, include_adm0_line = TRUE) 



```

Save the plot:

```{r saving_plot, eval = FALSE}
who_map_save("who_region_map.png", dpi = 700)
```


## Merging data to plot:

As stated previously, data can be merged with shapefiles to plot additional variables:

```{r vaccine_plot }

vaccine_sf <- left_join(
  sfs$adm0,
  vaccine_data %>% select(-report_country, -who_region),
  by = c("iso_3_code" = "iso3")
) %>% 
  mutate(persons_fully_vaccinated_per100 = 
           ifelse(persons_fully_vaccinated_per100 > 100, 100, persons_fully_vaccinated_per100))


ggplot() +
  geom_sf_who_poly(data = vaccine_sf, aes(fill = persons_fully_vaccinated_per100)) +
  scale_fill_fermenter(name = "Fully vaccinated per 100", 
                       palette = "RdYlGn", limits = c(0, 100),
                       direction = 1, na.value = who_map_col("no_data")) +
  labs(title ="Persons fully vaccined per 100, COVID-19", 
       subtitle = str_glue("as of {format(max(vaccine_sf$date_updated, na.rm = TRUE), '%d %b %y')}")) +
  who_map_pipeline(sf = sfs) 



```


## Producing regional maps  

It is also possible to show produce maps at country or regional level, based on changing the boundaries of the map. For example, to show the same map in the context of SEARO, we can edit the shapefiles to not show data for other countries. We can then zoom into the specified area, either by supplying coordinates, or generating a bounding box and suppling it to `who_map_pipeline()`.

Use of alternative WHO logos is also possible, via the region option in `who_map_pipeline()`

```{r searo map}

vaccine_sf_searo <- vaccine_sf %>% 
  mutate(persons_fully_vaccinated_per100 = ifelse(who_region == "SEARO", persons_fully_vaccinated_per100, NA_real_))

searo_box <- sf::st_bbox(vaccine_sf_searo %>% 
                           filter(who_region == "SEARO"))



ggplot() +
  geom_sf_who_poly(data = vaccine_sf_searo, aes(fill = persons_fully_vaccinated_per100)) +
  scale_fill_fermenter(name = "Fully vaccinated per 100", 
                       palette = "RdYlGn", limits = c(0, 100),
                       direction = 1, na.value = who_map_col("no_data")) +
  labs(title ="Persons fully vaccined per 100, COVID-19", 
       subtitle = str_glue("as of {format(max(vaccine_sf$date_updated, na.rm = TRUE), '%d %b %y')}")) +
  who_map_pipeline(sfs, box_lims = searo_box, region = "searo", include_adm0_line = TRUE) 

```

<br> 
<br>

If needs require, it is also possible to change map projections. Note that in doing this, the coordinates of maps are altered, and will need to be controlled for when setting limits. The use of the `bom_lims` argument of the `who_map_pipeline()` function is effective for this. For example, for a WHO EURO map:

```{r euro map}

sfs_reproj <- map(sfs, 
                  ~mutate(.x, geometry = st_transform(geometry, "+proj=laea +lon_0=39.8 +lat_0=34.7")))

box <- sfs_reproj$adm0 %>% 
  filter(who_region == "EURO") %>% 
  st_bbox()


ggplot() +
  geom_sf_who_poly(data = sfs_reproj$adm0, aes(fill = who_region)) +
  scale_fill_brewer(palette = "Set3", 
                    na.value = who_map_col("no_data"),
                    name = "WHO Region") +
  labs(title ="World map by WHO region", 
       subtitle = str_glue("as of {format(Sys.Date(), '%d %b %y')}")) +
  who_map_pipeline(sf = sfs_reproj, box_lims = box)


```

Different projections are especially useful when displaying specific areas of the world - a number of projections can be found at [spatialreference.org](spatialreference.org) - the user can search for a specific file and display it in proj4 format.

## Plotting country maps

We can also use whotools to produce country-specific maps. Note that these maps map require significantly more work to produce, and may require some adjustment to work with the WHO themes.

We can use the example of Indonesia to demonstrate, simulating arbitrary risk classification data to to highlight administrative regions. Note that it may help to use to pass a `col` argument to `geom_sf_who_poly()` to highlight borders - this is often not needed for the country level, as a separate line shapefile is supplied:


```{r idn map}

## pull in shapefiles for indonesia

sf_idn <- pull_sfs(adm_level = 1, iso3 = "IDN", query_server = FALSE)


## reproject files to centre on indonesia
projection <- "+proj=longlat +a=6378160 +b=6356774.50408554 +no_defs "
sf_idn <- sf_idn %>% 
  mutate(geometry = st_transform(geometry, projection))

## create bounding box around plotting area to define limits
bbox  <- sf_idn %>% 
  st_bbox()

## simulate random risk category data
categories <- c("High", "Medium", "Low")

sf_idn <- sf_idn %>% 
  mutate(category = factor(sample(categories, replace = TRUE, size = n())))


## plot data
ggplot() +
  geom_sf_who_poly(data = sf_idn, aes(fill = category),
                   col = who_map_col("border")) +
  scale_fill_manual(values = c("High" = "#fc8d59", "Medium" = "#ffffbf", "Low" = "#91cf60"),
                     breaks = c("High", "Medium", "Low")) +
  labs(title ="Indonesia mapping example",
       subtitle = str_glue("as of {format(Sys.Date(), '%d %b %y')}")) +
  who_map_pipeline(sf = sf_idn, box_lims = bbox, include_adm0_line = FALSE)


```

We can also pull in global shapefiles for admin 0 level to provide a background to the plot. This is especially helpful when it comes to Indonesia, as some borders bisect islands. This may be especially important when dealing with countries that have disputed areas adjacent to them. It should be noted that this solution uses all adm0 countries - to speed up plotting, the user can filter out countries not in the vicinity of the area of interest. This can be done as in the following:

```{r idn map adjustment}

## reproject global data 
sfs_global_idn <- map(sfs,
                      ~mutate(.x, geometry = st_transform(geometry, "+proj=longlat +a=6378160 +b=6356774.50408554 +no_defs")))


## plot data
ggplot() +
  ## first plot background adm0 polygons
  geom_sf_who_poly(data = sfs_global_idn$adm0, fill = who_map_col("no_data"),
                   col = who_map_col("border")) +
  ## plot as normal
  geom_sf_who_poly(data = sf_idn, aes(fill = category), col = who_map_col("border")) +
  scale_fill_manual(values = c("High" = "#fc8d59", "Medium" = "#ffffbf", "Low" = "#91cf60"),
                    breaks = c("High", "Medium", "Low")) +
  labs(title ="Indonesia mapping example",
       subtitle = str_glue("as of {format(Sys.Date(), '%d %b %y')}")) +
  ## change out sfs for global sfs
  who_map_pipeline(sf = sfs_global_idn, box_lims = bbox, include_adm0_line = TRUE)


```


We can finish this map off my adding labels to administrative regions. Doing this for every region would create a very busy plot, so we'll restrict this to regions with a classification level of _high_. There is no specific packaged function for this, as there are no specific WHO defaults to control for. This can be done as in the following:

```{r idn map final}

## reproject global data 
sfs_global_idn <- map(sfs,
                      ~mutate(.x, geometry = st_transform(geometry, "+proj=longlat +a=6378160 +b=6356774.50408554 +no_defs")))


## plot data
ggplot() +
  ## first plot background adm0 polygons
  geom_sf_who_poly(data = sfs_global_idn$adm0, fill = who_map_col("no_data"),
                   col = who_map_col("border")) +
  ## plot as normal
  geom_sf_who_poly(data = sf_idn, aes(fill = category), col = who_map_col("border")) +
  
  geom_sf_text(data = sf_idn %>% filter(category == "High"), 
               aes(label = str_wrap(adm1_viz_n, 10)),
               color = "#48494B",size=3, check_overlap = TRUE) +

  scale_fill_manual(values = c("High" = "#fc8d59", "Medium" = "#ffffbf", "Low" = "#91cf60"),
                    breaks = c("High", "Medium", "Low")) +
  labs(title ="Indonesia mapping example",
       subtitle = str_glue("as of {format(Sys.Date(), '%d %b %y')}")) +
  ## change out sfs for global sfs
  who_map_pipeline(sf = sfs_global_idn, box_lims = bbox, include_adm0_line = TRUE)


```
